### 2018-05-26 1_intro.tex
Cleanup finale up to Chap 3.

%TODO: rename mousquetaires: http://www.cprover.org/wmm/musketeer/ !!! (+reference this project: "Don’t sit on the fence".pdf in alglave/ )
% TODO: see how musketeer deals with recursive calls: http://www.cprover.org/wmm/musketeer/rec.pdf

@@

As an example, consider a shared memory concurrent system that executes two parallel processes as it is described in Figure~\ref{simple_wmm_x86} (such little examples that illustrate specific behaviour of a concurrent execution environment are called \textit{litmus tests}).
The process~\texttt{P0} writes the value~\texttt{1} to the shared variable~\texttt{x} and reads a value of the shared variable~\texttt{y}, and the process~\texttt{P1} writes the~\texttt{y} and reads the~\texttt{x}.
Considering all interleavings of the processes, one could expect the final state to be one of the following:
\begin{itemize}[noitemsep,topsep=0pt]
\item \texttt{(0:EAX=0,\,1:EBX=1)},
\item \texttt{(0:EAX=1,\,1:EBX=0)},
\item \texttt{(0:EAX=1,\,1:EBX=1)}.
\end{itemize}
%TODO!!!!!!!!!!!!!!!!!!! cite "mcKenney - Memory Barriers a Hardware View for Software Hackers.pdf" -- a table for multiple architectures

%As the program specifies the write to~\texttt{x} to precede the read of~\texttt{y} in process~\texttt{P0}, one could expect the final state `\texttt{(0:EAX=0~/\textbackslash~1:EAX=0)}' to be unreachable.
However, on the x86 architecture the state `\texttt{(0:EAX=0,\,1:EAX=0)}' is also reachable as the processors may cache the write to shared memory into their local write buffers, so that the write does not immediately become visible by processes running on other cores.
This behaviour is known as \textit{Store Buffering~(SB)}.

\begin{figure}[t]
\centering
\ttfamily
\begin{tabular}{ |l|l| }
\hline
\multicolumn{2}{|l|}{ \{ x=0; y=0; \}} \tabularnewline \hline
P0 & P1 \\ \hline
MOV [x],1 & MOV [y],1 \\
MOV EAX,[y] & MOV EBX,[x] \\
\hline
%\multicolumn{2}{|l|}{locations [x;y;]} \tabularnewline
\multicolumn{2}{|l|}{exists (0:EAX=0,~1:EBX=0)} \tabularnewline
\hline
\multicolumn{2}{|l|}{x86: allow} \tabularnewline
\hline
\end{tabular}
\caption{Store buffering: A litmus test illustrating the write-read reordering allowed by the x86-TSO memory model}
\label{simple_wmm_x86}
\end{figure}

@@

The SC model is considered to be a \textit{strong memory model} in the sense that it provides firm guarantees regarding the ordering and effect of memory operations.
%However, execution environments that provide weakened guarantees (\textit{relaxations} such as memory operations reordering, write buffering, etc.)
Weakened guarantees (such as memory operations reordering, write buffering, etc.) are called \textit{relaxations} of the memory model.
The relaxations of the SC model lead to \textit{Weak Memory Models~(WMMs)} that specify how processes interact through the shared memory, when a write becomes visible to processes running on other cores, and what value a read operation can get.
Thus, WMMs serve as set of guarantees made by designers of an execution environment to programmers on which behaviours of their concurrent code they can rely on.

@@

%TODO: litmus test synthesis 
%1) http://research.nvidia.com/sites/default/files/pubs/2017-04_Automated-Synthesis-of//ASPLOS_2017_Memory_Model_Verification.pdf

@@

Furthermore, there exists a wide range of tools that perform memory model-aware analysis.
\begin{itemize}[noitemsep,topsep=0pt]

@@

\item There exist tools that perform the weak memory model-aware program verification and model checking.
The notable examples are
%the tool \tool{mmchecker} for model-checking the C\# code~\cite{huynh2006memory}
the stateless model checkers \tool{RCMC}~\cite{kokologiannakis2017effective}, \tool{CHESS}~\cite{musuvathi2008fair} and \tool{Nidhugg}~\cite{abdulla2017stateless},
the tool \tool{Trencher} for checking programs against the TSO memory model~\cite{bouajjani2013checking}, 
the tool \porthos{} for portability analysis~\cite{Porthos17a},

@@

\item Some other tools perform static instrumentation of concurrent C programs and encode the WMM into the program representation so that it can be model-checked by standard tools.
The examples are
%TODO: cite the 'wmm-general/Thread-Modular Static Analysis for Relaxed Memory Models.pdf'
the instrumenting compiler \tool{goto-cc} which is a part of \tool{CBMC} model checker~\cite{kroening2014cbmc},
the tool that performs the sequentialisation of concurrent programs~\cite{alglave2013software},
the tool \tool{Weak2SC} for producing program descriptions which can be fed into standard model checking tools (such as \tool{SPIN}~\cite{holzmann1997model} or \tool{NuSMV}~\cite{cimatti2000nusmv}) for performing memory model-aware analysis~\cite{travkin2016verification}.
\end{itemize}

@@

All the tools listed above consider only a single memory model, however, in real life we face serious engineering problems involving necessity to model more than one execution environment.
One of these problems is the \textit{portability} of the program from one hardware architecture to another.
A program written in a high-level language is then compiled for different hardware.
Even if all the compiler optimisations were disabled (which is rare case nowadays), the behaviour of two compiled versions of the same program may differ due to differences between hardware memory models.
As the result, a program compiled under the platform $\mathcal{T}$ can reach states that are unreachable on the platform $\mathcal{S}$, which is a \textit{portability bug} from the source platform $\mathcal{S}$ to the target platform $\mathcal{T}$~\cite{Porthos17a}.

The very first tool that performs the WMM-aware portability analysis is \porthos{}%
%
\footnote{The \porthos{} project web site: \url{http://github.com/hernanponcedeleon/Dat3M}} %
%
introduced in April~2017~\cite{Porthos17b}.
This tool reduces described problem to a bounded reachability problem, which can be solved via an SMT solver.
This approach allows to capture symbolically the semantics of analysing program and both weak memory models into a single SMT-formula, augmented by the reachability assertion.
As most modern SMT solvers are efficient enough to be able to operate the state space of size millions of variables bounded by millions of constraints~(\cite{malik2009boolean}), the used method can be applicable in solving the real-world problems.
% TODO: search space size: https://courses.cs.washington.edu/courses/csep573/11wi/lectures/ashish-satsolvers.pdf slide

>>>>

%TODO: rename mousquetaires: http://www.cprover.org/wmm/musketeer/ !!! (+reference this project: "Don’t sit on the fence".pdf in alglave/ )

@@

As an example, consider two x86 assembly programs in Figure~\ref{simple_wmm_x86} (such little code examples that explain specific behaviour of the concurrent execution environment are called \textit{litmus tests}).
Two programs are the same, except the right-hand program uses the synchronisation instruction \texttt{MFENCE} after the store to the shared memory.
In both programs, the process~\texttt{P0} writes the value~\texttt{1} to the shared variable~\texttt{x} and reads a value of the shared variable~\texttt{y}, and the process~\texttt{P1} writes to the~\texttt{y} and reads the~\texttt{x}.

\begin{figure}[t]
\centering
\ttfamily
\begin{subfigure}[b]{.4\textwidth}
\begin{tabular}[b]{ |l|l| }
\hline
\multicolumn{2}{|c|}{ \{ x=0; y=0; \}} \\ \hline
\multicolumn{1}{|c|}{P0} & \multicolumn{1}{c|}{P1} \\ \hline
MOV [x],1 & MOV [y],1 \\
MOV EAX,[y] & MOV EBX,[x] \\
& \\
\hline
\multicolumn{2}{|l|}{exists (0:EAX=0,\,1:EBX=0)} \tabularnewline
\hline
\multicolumn{2}{|l|}{x86-TSO: allowed} \tabularnewline
\hline
\end{tabular}
\caption{Without synchronisation}
\label{simple_wmm_x86:nosync}
\end{subfigure}
\hfill
%
\begin{subfigure}[b]{.4\textwidth}
\begin{tabular}{ |l|l| }
\hline
\multicolumn{2}{|c|}{ \{ x=0; y=0; \}} \\ \hline
\multicolumn{1}{|c|}{P0} & \multicolumn{1}{c|}{P1} \\ \hline
MOV [x],1 & MOV [y],1 \\
MFENCE & MFENCE \\
MOV EAX,[y] & MOV EBX,[x] \\
\hline
\multicolumn{2}{|l|}{exists (0:EAX=0,\,1:EBX=0)} \tabularnewline
\hline
\multicolumn{2}{|l|}{x86-TSO: forbidden} \tabularnewline
\hline
\end{tabular}
\caption{With synchronisation}
\label{simple_wmm_x86:sync}
\end{subfigure}
%
\caption{Store buffering~(SB): A litmus test illustrating the write-read reordering allowed by the x86-TSO memory model}
\label{simple_wmm_x86}
\end{figure}

Taking into account all possible interleavings of the read and write instructions, both litmus tests allow the following final states:
\begin{itemize}[noitemsep,topsep=0pt]
\item \texttt{(0:EAX=0,~1:EBX=1)},
\item \texttt{(0:EAX=1,~1:EBX=0)},
\item \texttt{(0:EAX=1,~1:EBX=1)}.
\end{itemize}
%As the program specifies the write to~\texttt{x} to precede the read of~\texttt{y} in process~\texttt{P0}, one could expect the final state `\texttt{(0:EAX=0~/\textbackslash~1:EAX=0)}' to be unreachable.
However, in the left-hand (non-synchronised) litmus test, the state `\texttt{(0:EAX=0,~1:EAX=0)}' is also allowed on the x86 architecture as its processors may cache the writes to the shared memory into their local \textit{store buffers} so that the updated value does not immediately become visible by processes running on other cores.
This behaviour is known as \textit{Store Buffering~(SB)}.
For preventing such a behaviour, the x86 architecture offers the synchronisation instruction \texttt{MFENCE} that flushes the store buffer of the process.
In Figure~\ref{simple_wmm_x86:sync}, this instruction is inserted for both processes after the store instructions, thus once it is executed in one process, the other process will read the updated value of this shared memory location.
The work~\cite{mckenney2010memory} gives a comprehensive review of common hardware architectures from the perspective of the structure of their caches and write buffers, that determine the effect of memory operations, and memory barriers they provide.

@@

The SC model is considered to be a \textit{strong memory model} in the sense that it provides firm guarantees regarding the ordering and effect of memory operations.
%However, execution environments that provide weakened guarantees (\textit{relaxations} such as memory operations reordering, write buffering, etc.)
Weakening of the guarantees (such as memory operations reordering, write buffering, etc.) is called \textit{relaxation} of the memory model.
The relaxations of the SC model lead to \textit{Weak Memory Models~(WMMs)} that specify how processes interact through the shared memory, when a write becomes visible to processes running on other cores, and what value a read operation can get.
Thus, WMMs serve as a set of guarantees made by designers of an execution environment to programmers on which behaviours of their concurrent code they can rely on.

@@

@@

Furthermore, there exists a wide range of tools that perform memory model-aware analysis.

\begin{itemize}[noitemsep,topsep=0pt,leftmargin=\parindent]

@@

\item Some tools that perform the weak memory model-aware program verification and model checking.
The notable examples are
%the tool \tool{mmchecker} for model-checking the C\# code~\cite{huynh2006memory}
the stateless model checkers \tool{RCMC}~\cite{kokologiannakis2017effective}, \tool{CHESS}~\cite{musuvathi2008fair}, \tool{Nidhugg}~\cite{abdulla2017stateless}, \tool{Trencher}~\cite{bouajjani2013checking}, and others.

@@

\item Along with synthesis of synchronisation primitives, some tools were designed to perform the automatic synthesis of litmus tests for the weak memory model, for example \tool{litmustestgen} by the NVidia Research~\cite{lustig2017automated}.

\item The framework \tool{Alloy}~\cite{wickerson2017automatically} can be used for understanding WMMs; it can automatically generate conformance tests between two memory models, distinguish two WMMs, and check monotonicity and compiler mappings of a program.

\item Some other tools perform static instrumentation of concurrent C programs and encode the WMM into the program representation so that it can be model-checked by standard tools.
The examples are:
\begin{itemize}[noitemsep,topsep=0pt]
\item the instrumenting compiler \tool{goto-cc} which is a part of \tool{CBMC} model checker~\cite{kroening2014cbmc},
\item the tool that performs the sequentialisation of concurrent programs~\cite{alglave2013software},
\item the tool \tool{Weak2SC} generates weak memory model descriptions of the program, which can be fed into standard model checking tools (such as \tool{SPIN}~\cite{holzmann1997model} or \tool{NuSMV}~\cite{cimatti2000nusmv}) for performing memory model-aware analysis~\cite{travkin2016verification}.
\end{itemize}

\end{itemize}

@@

All the tools listed above consider only a single memory model, however, in real life we face serious engineering problems involving more than one execution environment.
One of these problems is the \textit{portability analysis} of the program from one hardware architecture to another.
A program written in a high-level language is compiled for different hardware architectures.
Even if all the compiler optimisations were disabled (which is a rare case nowadays), the behaviour of two compiled versions of the same program may differ due to differences between hardware memory models.
As the result, a program compiled under the platform $\mathcal{T}$ can reach states that are unreachable on the platform $\mathcal{S}$, which is a \textit{portability bug} from the source platform $\mathcal{S}$ to the target platform $\mathcal{T}$~\cite{Porthos17a}.

The very first tool that performs the WMM-aware portability analysis is \porthos{}%
%
\footnote{The \porthos{} project web site: \url{http://github.com/hernanponcedeleon/PORTHOS}} %
%
introduced in April~2017~\cite{Porthos17a}.
This tool reduces described problem to a bounded reachability problem, which can be solved via an SMT solver.
This approach allows to capture symbolically the semantics of analysing program and both weak memory models into a single SMT-formula, augmented by the reachability assertion.
As most modern SMT solvers are efficient enough to be able to operate the state space of size millions of variables bounded by millions of constraints~(\cite{malik2009boolean}), the used method can be applicable in solving the real-world problems.
% about SAT: https://courses.cs.washington.edu/courses/csep573/11wi/lectures/ashish-satsolvers.pdf

### 2018-05-26 2_wmm.tex
Cleanup finale up to Chap 3.

One way to fight the combinatorial explosion is to exclude the global time from the model and treat executions from one equivalence class together in a non-deterministic fashion.
For instance, such an equivalence class can be the set of computations performed by a processor locally that do not affect the global state.
This idea is used in the \textit{event-based} model, that represents the program as a directed graph of events (the \textit{event-flow graph})~\cite{alglave2010shared}.
The vertices of such a graph represent \textit{events} (see Section~\ref{ch:wmm:model:events}), and edges represent basic relations (see Section~\ref{ch:wmm:model:relations}).
The graph represents the set of executions (sequences of events; see Section~\ref{ch:wmm:model:executions}) defined by the non-deterministic guesses of certain relations on some states.

@@

%- The semantics of the event-flow graph is given:  "alglave: Don’t sit on the fence A static analysis approach to automatic fence insertion" page 8 or 9
%"To build the aeg , we define a semantics of goto-programs in terms of abstract events, ..."
% TODO: OR from here: http://www.cprover.org/etaps/esop.pdf "Software Verification for Weak Memory via Program Transformation"

@@

%TODO: EVENT -- ATOMIC?
An event is a fact of executing the low-level primitive operation such as memory access, threads synchronisation, computation, etc.

@@

The synchronisation instructions (fences) cause the \textit{barrier events}, that do not perform any computation or memory value transfer, instead, they add new relations to the program model that restrict the set of allowed behaviours.
Functionally, a fence may be a synchronisation barrier or a instruction of flushing the local memory caches, etc.

% TODO: more fences: "On Power and ARM (see Sec. 6), we distinguish between control fence (which we write cfence), lightweight fence (lwfence) and full fence (ffence). On x86, implementing TSO, there is only one fence, called mfence." (herding cats, p. 14)

@@

\vspace{1em}
The basic relations are the following~\cite{alglave2010shared}:
%TODO: say that po and rf are  defined from the candidate execution:
%TODO: say that 
\begin{itemize}
    \item The \textit{control-flow} of a program is defined by the \textit{program-order} relation \po~$\subseteq~\mathbb{E}~\times~\mathbb{E}$, which represents the total order of events of same process.
    For instance, if the instruction $i_1$ generates the event $e_1$ and the instruction $i_2$ follows $i_1$ and generates the event $e_2$, then $e_1 \relarrow{po} e_2$.
    %Some new relations may be acquired : dp, po-loc

@@

%TODO: add ppo = characteristics of memory model. "Architectures are instances of our model. An architecture is a triple of functions (ppo, fences, prop), which specifies the preserved program order ppo, the fences fences and the propagation order prop." (herding cats, p.13)

@@

    %TODO: say that po-relation is *static*
    \item the \texttt{po-loc} relation that is the \po-relation over events that access to the same shared variable: \\
    $m_1 \relarrow{po-loc} m_2 ~\triangleq~ (m_1 \relarrow{po} m_2 \land \texttt{loc}(m_1) = \texttt{loc}(m_2))$; and

@@

%todo: some words here?

@@

The semantics of a concurrent program is represented as the set of allowed executions.
The \textit{execution} is a path in the event-flow graph defined by \po- and \rf-relations and set of final writes to a given memory location that is valid under certain memory model~\cite{alglave2014herding}.
It can be interpreted as a sequence of guesses which event is to be executed next.
The \textit{candidate execution} is an execution that is not yet constrained by a memory model.

%TODO: cite this paper elsewhere
%As it was shown in~\cite{wickerson2017automatically}, it is enough for memory models to constrain the executions independently instead of constraining the program at a whole.

%An \textit{execution} (trace, run) of a program is an ordered set of events defined by \po- and \rf-relations and set of final writes to a given memory location that is valid under certain memory model~\cite{alglave2014herding}.
%The order of events in particular execution is denoted as `$\rightarrow$', an empty execution is denoted as $\emptyset$.
%associated with the instructions of the program
%An execution is considered to be \textit{valid} if the memory events follow a single global timeline, i.e., can be embedded in a single partial order allowed by the memory model restrictions~\cite{alglave2010shared}.
%An execution is uniquely defined by the set $\mathbb{X}$ of events have been executed in each thread (the \textit{control-flow} of a program), and the relations $\texttt{rf}$ and $\texttt{co}$~\cite{alglave2010shared}.
Figure~\ref{simple_wmm_x86_pic} illustrates four possible candidate executions for the litmus test Example~\ref{simple_wmm_x86} (the pictures are generated by the \tool{herd7} tool, version~7.47).
Since there are no conditional jumps, the \po-relation is defined and we do not need to guess it.
Since each thread performs single write followed by a single read, the \co-relation is also defined (it relates the initial write event with the write event to the same location).

@@

Weak memory models are defined via \cat{} language~\cite{alglave2016syntax}.
This is a domain specific language for describing consistency properties of concurrent programs.
The language combines expressive power of a functional language (it is inspired by OCaml and adopts its types, first-class functions, pattern matching and other features) with types, operations and assertions that are specific for operating with relations and executions.
%represents the functional language extended by the theory of  allows to axiomatically define derived relations, architecture-specific fences, and constraints over relations.
In \cat{}, new relations can be defined via the keyword \texttt{let} and the following operators over relations~\cite{alglave2016syntax}. % 3.1.6 Relations between events

%TODO: this is binary relations. ADD UNARY RELS!

Below we enumerate pre-defined operators over relations and sets of events:

@@

For instance, the \fr-relation is defined as a sequence of inverted \rf-relation and \co-relation: $\texttt{fr} = (\invrel{rf};\texttt{co})$.
As an example of memory model definition in \cat{} language, Figure~\ref{ex:x86-cat} presents the simplified x86-TSO model~\cite{herd10tutorial}.
This memory model asserts  acyclicity of the communication relation (the union of \rf-, \fr- and \co- relations), \texttt{po-loc} -relation, \texttt{mfence}-relation and some other derived relations~\cite{owens2009better}.

\begin{figure}[h]
\begin{lstlisting}
"X86 TSO"
include "x86fences.cat"
include "filters.cat"
include "cos.cat"

(* Uniproc check *)
let com = rf | fr | co
acyclic po-loc | com

(* Atomic *)
empty rmw & (fre;coe)

(* Global happens-before *)
#ppo
let po_ghb = WW(po) | RM(po)

#mfence
include "x86fences.cat"

#implied barriers
let poWR = WR(po)
let i1 = MA(poWR)
let i2 = AM(poWR)
let implied = i1 | i2

let com = rfe | fr | co
let ghb = mfence | implied | po_ghb | com
show implied

acyclic ghb as tso
\end{lstlisting}
\caption{The simplified x86-TSO memory model in the \cat{} language~\cite{herd10tutorial}}
\label{ex:x86-cat}
\end{figure}

>>>>

One way to fight the combinatorial explosion is to exclude the global time from the model and treat executions from one equivalence class together in a non-deterministic fashion.
For instance, such an equivalence class can be the set of computations performed by a processor locally that do not affect the global state.
This idea is used in the \textit{event-based} model, that represents the program as a directed graph of events (the \textit{event-flow graph})~\cite{alglave2010shared,alglave2014herding}.
The vertices of such a graph represent \textit{events} (see Section~\ref{ch:wmm:model:events}), and edges represent basic relations (see Section~\ref{ch:wmm:model:relations}).
The graph represents the set of executions (sequences of events; see Section~\ref{ch:wmm:model:executions}) defined by the non-deterministic guesses of certain relations on some states.

@@

@@

An \textit{event} is a fact of executing the low-level primitive atomic operation such as memory access, threads synchronisation, computation over the local-memory, control-flow jump, etc.

@@

Synchronisation instructions (fences) cause \textit{barrier events}, which do not perform any computation or memory value transfer, instead, they the set of program behaviours by adding barrier relations to the program model.
Functionally, a fence may be a synchronisation barrier or a instruction for flushing memory caches into the main memory, etc.
For instance, the \texttt{mfence} instruction of the x86 assembly flushes the store buffers of the thread, and thus does not allow the \rf-relation (see Section~\ref{ch:wmm:model:relations}) to point to

@@

\vspace{1em}
The basic relations are the following~\cite{alglave2010shared}:
\begin{itemize}
    \item The \textit{control-flow} of a program is defined by the \textit{program-order} relation \po~$\subseteq~\mathbb{E}~\times~\mathbb{E}$, which represents the total order of events of same process.
    For instance, if the instruction $i_1$ generates the event $e_1$ and the instruction $i_2$ follows $i_1$ and generates the event $e_2$, then $e_1 \relarrow{po} e_2$.
    %Some new relations may be acquired : dp, po-loc

@@

%TO-DO: add ppo = characteristics of memory model. "Architectures are instances of our model. An architecture is a triple of functions (ppo, fences, prop), which specifies the preserved program order ppo, the fences fences and the propagation order prop." (herding cats, p.13)

@@

    %TO-DO: say that po-relation is *static*
    \item the \texttt{po-loc} relation that is the \po-relation over events that access to the same shared variable: \\
    $m_1 \relarrow{po-loc} m_2 ~\triangleq~ (m_1 \relarrow{po} m_2 \land \texttt{loc}(m_1) = \texttt{loc}(m_2))$; and

@@

@@

The semantics of a concurrent program is represented as the set of allowed executions.
An \textit{execution} is a path in the event-flow graph defined by \po- and \rf-relations and set of final writes to a given memory location that is valid under certain memory model~\cite{alglave2014herding}.
It can be interpreted as a sequence of guesses which event is to be executed next.
A \textit{candidate execution} is an execution that is not yet constrained by a memory model.

%An \textit{execution} (trace, run) of a program is an ordered set of events defined by \po- and \rf-relations and set of final writes to a given memory location that is valid under certain memory model~\cite{alglave2014herding}.
Figure~\ref{simple_wmm_x86_pic} illustrates four possible candidate executions for the litmus test Example~\ref{simple_wmm_x86} (the pictures are generated by the \tool{herd7} tool, version~7.47).
Since there are no conditional jumps, the \po-relation is defined and we do not need to guess it.
Since each thread performs a single write followed by a single read, the \co-relation is also defined (it relates the initial write event with the write event to the same location).

@@

Weak memory models are defined via \cat{} language~\cite{alglave2016syntax}.
It is a domain specific language for describing consistency properties of concurrent programs.
The language combines expressive power of a functional language (being inspired by OCaml, it adopts the OCaml types, first-class functions, pattern matching and some other features) with concepts of memory models (sets of events, relations, operations and assertions over relations).
In \cat{}, new relations can be defined via the keyword \texttt{let} and the following operators over relations~\cite{alglave2016syntax}.

Below we enumerate pre-defined operations over relations and sets of events:

@@

For instance, the \fr-relation is defined as a sequence of inverted \rf-relation and \co-relation: $\texttt{fr} = (\invrel{rf};\texttt{co})$.
As an example of memory model definition in \cat{} language, Figure~\ref{ex:x86-cat} presents the excerpt from the x86-TSO memory model~\cite{herd10tutorial}.
This memory model specification asserts  acyclicity of the communication relation (the union of \rf-, \fr- and \co- relations), \texttt{po-loc} -relation, \texttt{mfence}-relation and some other derived relations~\cite{owens2009better}.

\begin{figure}[H]
\begin{lstlisting}
...
let po_ghb = WW(po) | RM(po)
let implied = PA(poWR) | WR(po)
let GHB = mfence | implied | po_ghb | rfe | fr | co
let com = rf | fr | co

empty atom & (fre;coe)
acyclic po-loc | com
acyclic GHB
\end{lstlisting}
\caption{Excerpt from the x86-TSO memory model in the \cat{} language}
\label{ex:x86-cat}
\end{figure}
